Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics

### 2015-11-10 paper.tex
First half of references

@@

Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc. These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc. However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc. These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.  Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases. This has led to approaches such a ``micro-batching'' [], where a batch processor is used to simulate continuous processing, and the ``lambda architecture'', where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.

@@

Apache Flink is a platform that was designed to deliver on these requirements. It is an open source project licensed by the Apache Software Foundation (ASF), with its community counting more than 120 developers, and growing fast []. Flink originates from the database research community, in particular the Stratosphere research project [], and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages. In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.

This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results. Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high. Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions [socc10, vldb12] and efficiently embed iterations within a DAG-based dataflow system [vldb12].

@@

\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs. A Job Graph is a DAG of operators and connections between operators. There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing''). Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine. On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs. Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.

A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager. The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager. The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types. DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).

@@

Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals. When failure occurs the state of the execution is restored from the latest snapshot. Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source. Normally this possible by using persistent message queueing systems, e.g. Kafka.

Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS) [arXiv paper]. It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner. It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream. Similar to Chandy-Lamport distributed snapshots [Chandy/Lamport], the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum. ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further. Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records. Furthermore, ABS [arXiv paper] supports snapshotting of cyclic graphs.  The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated. The following figure illustrates the snapshot process.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ABS figure comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In distributed systems, there is an arbitrary lag between event-time and processing-time [dataflow]. This may mean arbitrary delays for getting an answer based on event-time semantics. To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system. The watermarks aid the execution engine to process events in the correct event order.

@@

\textbf{Streaming Windows.} A window defines a logical group of records that are processed together. Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google. A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor. 
The assigner predefines the logical groups to which each record belongs, e.g. count-based or time-based periodic windows. The trigger defines when the window operation is performed on each group. The evictor defines which records to keep in each group. The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger). Flink also offer shart hands for commonly used functionalities.

@@

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality [Dataflow], and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).

@@

A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above. However, i) the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops) and ii) we can simplify the fault tolerance mechanisms and iii) we can apply query optimization borrowing ideas from MPP database systems.

For these reasons, Flink treats specially batch computations, by implementing the above optimizations. For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime. These operators simply block until they have received all of their input. Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad [Dryad].

@@

\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties. However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink's optimizer employs a number of novel methods for overcoming these issues [blackboxes, stratosphere-journal, iterations] for which we provide a short overview below. Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation [scope-optimizer], using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier,  Flink's optimizer uses hints that are provided by the programmer.

\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments. These memory segments resemble database blocks into which, java objects representing the tuples that go through the runtime are serialized. Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.

\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine. Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG [Hadoop iterative works, Spark], hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad [] adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Giraph, and GraphLab) allowing to reduce the number of computations in each iteration [].

Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory. This approach simulates a cyclic dataflow within a DAG engine making Flink i ) competitive with specialized graph engines [], and outperforms the driver-based approach []. Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API. Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

@@

\section{References}

>>>>

% url references
\usepackage{hyperref}

% in paragraph enums
\usepackage{paralist}

@@

Various real-time analytics and CEP tools are currently in use by the industry, including TIBCO, IBM Infosphere Streams, Microsoft Streaminsight, Streambase, etc. These systems implement and extend the results of research which has been done in the database community in the last 15 years, with projects such as Aurora, Telegraph, STREAMS, etc. However, the era of Web-scale computing, spawned the development of internal ``roll-your-own'', solutions based on systems like Apache Storm, Samza, Kafka, Spark, etc. These systems provide very low level APIs and enforce only custom, application-specific solutions to be built on top.  Moreover, in the lack of better solutions, the industry is using traditional batch processing platforms to serve streaming use cases. This has led to approaches such a ``micro-batching''~\cite{Zaharia:2013:DSF:2517349.2522737}, where a batch processor is used to simulate continuous processing, and the ``lambda architecture''~\cite{marz2015big}, where a stream processor is aided by a batch processor to compensate for lack of performance and consistency guarantees.

@@

Apache Flink is a platform that was designed to deliver on these requirements. It is an open source project licensed by the Apache Software Foundation (ASF)~\cite{CUSTOM:web/Apache}, with its community counting more than 120 developers, and growing fast.~\cite{CUSTOM:web/FlinkContrib} Flink originates from the database research community, in particular the Stratosphere research project~\cite{stratosphere}, and incorporates several novel research contributions, some of which have been published in research venues in the past, and some of which have only been published in the form of blog posts or wiki pages. In this article, we attempt a short description of the architecture of the system, highlighting its novel aspects.

This paper is organized as follows: Section X presents a novel paradigm for distributed dataflows based on the notions of operators and logical intermediate results. Section Z presents a novel mechanism to provide fault tolerance while keeping the latency low and the throughput high. Section Y presents methods to overcome limitations of the JVM by careful engineering, and how to while extend cost-based optimization to user-defined functions~\cite{battre2010nephele, DBLP:journals/pvldb/EwenTKM12} and efficiently embed iterations within a DAG-based dataflow system~\cite{DBLP:journals/pvldb/EwenTKM12}.

@@

\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs. A Job Graph is a DAG of operators and connections between operators. There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing''). Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine. On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs. Currently these are the following:
\begin{inparaenum}[i)]
  \item Gelly, an API and library for processing graphs,
  \item FlinkML, an API and library for composing Machine Learning pipelines and
  \item Table, an API similar in spirit to Microsoft's LINQ.
\end{inparaenum}
A Flink cluster comprises of three types of processes: the client, the JobManager, and the TaskManager. The client takes the program code (written in a mixture of Flink's APIs), transforms it to a Job Graph, and submits it to the JobManager. The compilation process involves a type extraction and checking phase that generates serializers and comparators for all used types. DataSet programs, also go through a cost-based query optimization phase, similar to the physical optimizations performed by relational query optimizers (more details in Section X).

@@

Fault tolerance in Flink is achieved by taking a snapshot of the execution graph at regular intervals. When failure occurs the state of the execution is restored from the latest snapshot. Upon recovery, in order to guarantee exactly-once processing semantics, the records (events) consumed since the latest snapshot are reprocessed in the same order at each respective source. Normally this possible by using persistent message queueing systems, e.g. Apache Kafka.~\cite{CUSTOM:web/Kafka}

Flink's core mechanism for distributed snapshots is called ``Asynchronous Barrier Snapshotting'' (ABS).~\cite{carbone2015lightweight} It creates a global snapshot by collecting the state of each task in the execution graph in an asynchronous manner. It does so by superimposing snapshotting using injected markers, called ``barriers'', in the input data stream. Similar to Chandy-Lamport distributed snapshots~\cite{chandy1985distributed}, the barriers in ABS dictate the records that should be fully processed before each respected global snapshot, however, no records in transit are included in a snapshot, keeping the persisted state at a minimum. ABS achieves this by a special ``aligning'' phase which ensures that all records from all upstream tasks preceding broadcasted barriers have been fully consumed before processing the data stream further. Recovery from failure reverts the execution graph to the most recent snapshot and is fully consistent by respecting the causal dependency of records. Furthermore, ABS~\cite{carbone2015lightweight} supports snapshotting of cyclic graphs.  The operation of ABS is fully decoupled from the backend used for persisting states, thus, allowing multiple different backend implementations to be integrated. The following figure illustrates the snapshot process.

\begin{figure}[h!]
	\centering  	
  	\includegraphics[width=.45\textwidth]{figs/snaps.pdf}
	\caption{Asynchronous Barrier Snapshotting.}
	\label{fig:FlinkStack}
\end{figure}

@@

In distributed systems, there is an arbitrary lag between event-time and processing-time.~\cite{akidau2015dataflow} This may mean arbitrary delays for getting an answer based on event-time semantics. To avoid arbitrary delays these systems regularly insert special events called ``low watermarks''  that include a time attribute $t$ indicating that all events lower than $t$ have already entered the system. The watermarks aid the execution engine to process events in the correct event order.

@@

\textbf{Streaming Windows.} A window defines a logical group of records that are processed together. Flink's windowing system largely follows the Dataflow model~\cite{akidau2015dataflow} proposed by Google. A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor. 
The assigner predefines the logical groups to which each record belongs, e.g. count-based or time-based periodic windows. The trigger defines when the window operation is performed on each group. The evictor defines which records to keep in each group. The following example defines windows that operate in event time of 6 seconds that slide every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger). Flink also offer short hands for commonly used functionalities.

@@

Note that streams are already partitioned on a key before windowing, so the above is a local operator that does not require coordination between machines. This mechanism can be used to implement a wide variety of windowing functionality~\cite{akidau2015dataflow}, and Flink comes bundled with syntactic sugar for the most common window definitions (e.g., time- and count-based windows).

@@

A bounded data set is a special case of an unbounded data stream. Thus, a streaming program that inserts all of its input data in a window can form a ``batch`` program and ``batch analytics``, or ``batch processing`` should be fully covered by Flink's features that we presented above. However, \begin{inparaenum}[i)]
  \item the syntax, i.e., the API for batch computation can be simplified (not need for window definitions, simpler joins and loops),
  \item we can simplify the fault tolerance mechanisms and
  \item we can apply query optimization borrowing ideas from MPP database systems.
\end{inparaenum}
For these reasons, Flink treats specially batch computations, by implementing the above optimizations. For enabling batch computations on top of it's streaming runtime, Flink embeds blocking versions of its operators (sorts, joins, etc) within its runtime. These operators simply block until they have received all of their input. Moreover, Flink currently disables the asynchronous snapshotting mechanism for batch programs, and simply uses backtracking based recovery as first described in Dryad.~\cite{isard2007dryad}

@@

\textbf{Query Optimization.} Flink's optimizer builds on techniques from parallel database systems such as plan equivalences, cost models and interesting properties. However, the arbitrary UDF-heavy DAGs that make up Flink's dataflow programs, do not allow a traditional optimizer to employ database techniques out of the box [black-boxes], since the operators hide their semantics from the optimizer. For the same reason, cardinality and cost estimation methods are equally difficult to employ. Flink's optimizer employs a number of novel methods for overcoming these issues~\cite{blackBoxes, stratosphere, DBLP:journals/pvldb/EwenTKM12} for which we provide a short overview below. Flink's runtime supports various execution strategies including repartition/broadcast data transfer, as well as sort-based grouping and sort- and hash-based join implementations. Flink's optimizer enumerates different physical plans based on the concept of interesting properties propagation~\cite{scopeOptimizer}, using a cost-based approach to choose among multiple physical plans. The cost includes network/disk I/O and CPU cost. To overcome the cardinality estimation issues that were mentioned earlier, Flink's optimizer uses hints that are provided by the programmer.

\textbf{Memory Management.} Building on database technology, Flink, instead of storing objects in the JVM's heap, serializes objects into a  memory segments. These memory segments resemble database blocks into which, Java objects representing the tuples that go through the runtime are serialized. Operations such as sorting, and joining, operate as much as possible on the binary data, keeping the de/serialization overhead at a minimum and partially spilling data to disk when needed. To handle arbitrary objects, Flink uses type inference, and  custom serialization mechanisms.  By keeping the data processing on binary representation and off-heap, Flink manages to reduce the garbage collection overhead, and implement cache-efficient and robust algorithms that scale gracefully in under memory constraints.

\textbf{Native Iterations on top of Dataflows.} The final aspect of Flink on which we focus, is how to implement loops on top of the Flink's dataflow engine. Some approaches execute iterations by submitting new jobs for each iteration or by adding additional nodes to a running DAG~\cite{DBLP:journals/pvldb/BuHBE10, DBLP:conf/hotcloud/ZahariaCFSS10}, hiding from  the engine that it is executing an iterative program. The approach, implemented in Naiad~\cite{murray2013naiad} adds feedback edges in the dataflow graph, supporting graphs with cycles, that  allow for nested iterations. A third approach was to design specialized engines around iterative processing along (e.g., Apache Giraph, and GraphLab) allowing to reduce the number of computations in each iteration.~\cite{low2012distributed}

Flink follows an approach that maintains the DAG-based runtime, but allows for special ``head'' and ``tail'' tasks to signify the beginning and end of iteration, that exchange data via shared memory. This approach simulates a cyclic dataflow within a DAG engine making Flink i) competitive with specialized graph engines [], and outperforms the driver-based approach []. Flink supports delta iterations [], which exploit sparse computational dependencies, and are used, as the basis for Gelly, Flink's Graph API. Finally, Flink offers Bulk Synchronous Parallel (BSP) iterations in its DataSet API, and asynchronous iterations in its DataStream API.

@@
